Aligning marine species range data to better serve science and conservation

### 2016-03-11 data_explore/data_explore.Rmd
updating manuscript, and creating maps for gaps and 10+% MPAs

>>>>

``` {r plot_spp_iucn_gaps_tenplus, eval = TRUE, echo = FALSE}

iucn_gap10     <- read_csv(file.path(dir_data, 'iucn_gaps_tenplus.csv'))
iucn_gap10_spp <- unique(iucn_gap10[iucn_gap10$iucn_sid %in% spp_list$iucn_sid, ])

iucn_gapmaps <- sample_n(iucn_gap10_spp %>% filter(cat == 'gap'), 10)
iucn_10maps  <- sample_n(iucn_gap10_spp %>% filter(cat != 'gap'), 10)

if(!exists('iucn_spp_cells')) {
  iucn_spp_cells <- read_csv(file.path(dir_anx, 'explore_am_v_iucn/iucn_spp_cells_2015.csv'),
                             col_types = 'cddddc')
  am_spp_cells   <- read_csv(file.path(dir_anx, 'explore_am_v_iucn/am_spp_cells_2015.csv'))
}

am_thresh = 0

message('Printing gap maps')
for (i in 1:nrow(iucn_gapmaps)) { ### i <- 1
  spp <- iucn_gapmaps[i, ]

  spp_plot <- plot_rangemap(spp)
  spp_plot <- spp_plot +
    labs(title = sprintf('%s: %s', spp$sciname, spp$spp_group_text))

  print(spp_plot)
}

message('Printing >10% maps')

for (i in 1:nrow(iucn_10maps)) { ### i <- 1
  spp <- iucn_10maps[i, ]
  spp_plot <- plot_rangemap(spp)
  spp_plot <- spp_plot +
    labs(title = sprintf('%s: %s', spp$sciname, spp$spp_group_text))

  print(spp_plot)
}

```
***

### 2016-03-11 manuscript-main.Rmd
updating manuscript, and creating maps for gaps and 10+% MPAs

---
title: "Aligning marine species range data to better serve science and conservation"
author: "Casey O'Hara^1^, Jamie Afflerbach^1^, Benjamin S. Halpern<sup>1, 2, 3</sup>"
output:
  html_document:
    highlight: haddock
    number_sections: yes
    theme: spacelab
    toc: no
  pdf_document:
    number_sections: yes
    toc: no
  word_document: default
---

>>>>

---
title: "Aligning marine species range data to better serve science and conservation"
author: Casey O'Hara^1^, Jamie Afflerbach^1^, Benjamin S. Halpern<sup>1, 2, 3</sup>,
  Courtney Scarborough^1^
output:
  word_document: default
  pdf_document:
    number_sections: yes
    toc: no
  html_document:
    highlight: haddock
    number_sections: yes
    theme: spacelab
    toc: no
---

### 2016-03-11 ms_2_intro_res_disc.Rmd
updating manuscript, and creating maps for gaps and 10+% MPAs

___BSH:___ _I think Julie makes this comment below, but we may need to be able to substantiate this somehow.  Is there a way to quickly figure out how many studies have used AM vs. IUCN?_

___CCO:___ _What about age - AquaMaps is pretty recent (2006 at earliest); before that, would IUCN have been the dominant dataset for marine?  Maybe we can track citations of AquaMaps, but citations of IUCN are generic - just cite the Red List - not referencing spatial information specifically, much less marine.  Going by citation year field on the maps themselves, the oldest marine record I can find is 2004._

_Three foundational papers relating to AquaMaps (Kaschner et al. 2006, 2011; Ready et al. 2010) have been cited a total of 315 times._

@@

 (<a href = '#fig5'>Figure 5</a>) displays the results of applying this methodology under new scenarios.  When comparing the conclusions based upon an IUCN-only scenario to an AquaMaps-only scenario (using 2015 AquaMaps data and a 0% threshold sets the most meaningful comparison) we found a five-fold increase in the proportion of gap species (6.4% of species vs. 1.2%) and dramatically larger proportion of species with less than 2% of their range protected (73.2% of species vs. 47.7%).  However, this comparison also indicates a larger proportion of well-protected species with greater than 10% of range protected (2.9% of species vs. 1.5%).  Conclusions drawn from each of these datasets would paint dramatically different pictures of the protection afforded by our current global MPA network.

___If you have room, I think you could add a bit more interpretation of these results here. Could discuss the commission and omission stuff here too to describe these somewhat non-intuitive results??? CS___

- check which species are gap species and look at some maps
- check which species are 10%+ species and look at some maps
- where do these species fall georegionally? is there anything odd about their ranges?

@@

<!-- * Additional filters on AquaMaps distribution criteria, such as expert review status or the number of occurrence cells used to generate the species model, can help avoid overextrapolation for data poor species. _using other attributes and variables to filter? this seems outside the scope of this paper_ -->

_(do we need a wrap-it-up paragraph here?  I put that more at the top of the conclusions section... could shift it down here, but probably not needed in both places)_

>>>>

___BSH comment:___ _I think Julie makes this comment below, but we may need to be able to substantiate this somehow.  Is there a way to quickly figure out how many studies have used AM vs. IUCN?_  ___CCO: I can give # of citations for AquaMaps foundational papers (~315); IUCN is tougher since IUCN citations aren't necessarily spatial or marine.  How important is this - i.e. can I get away without making a big deal here? or making a sweeping statement of some sort??___

<!--
___CCO:___ _What about age - AquaMaps is pretty recent (2006 at earliest); before that, would IUCN have been the dominant dataset for marine?  Maybe we can track citations of AquaMaps, but citations of IUCN are generic - just cite the Red List - not referencing spatial information specifically, much less marine.  Going by citation year field on the maps themselves, the oldest marine record I can find is 2004._

_Three foundational papers relating to AquaMaps (Kaschner et al. 2006, 2011; Ready et al. 2010) have been cited a total of 315 times._
-->

@@

 (<a href = '#fig5'>Figure 5</a>) displays the results of applying this methodology under new scenarios.  When comparing the conclusions based upon an IUCN-only scenario to an AquaMaps-only scenario (using 2015 AquaMaps data and a 0% threshold sets the most meaningful comparison to IUCN's "limits of distribution") we found a five-fold increase in the proportion of gap species (6.4% of species vs. 1.2%) and dramatically larger proportion of species with less than 2% of their range protected (73.2% of species vs. 47.7%).  However, this comparison also indicates a larger proportion of well-protected species with greater than 10% of range protected (2.9% of species vs. 1.5%).  Conclusions drawn from each of these datasets would paint dramatically different pictures of the protection afforded by our current global MPA network.

___working on checking why so many gaps and 10+ species show up with IUCN.  I will write up a paragraph once I've had a little more time to examine maps___

- check which species are gap species; check maps
- check which species are 10%+ species and check maps
- where do these species fall georegionally? is there anything odd about their ranges? any reason why IUCN would be 
- hypothesis to explain more gap species: Aquamaps more diffuse than IUCN, catching individual cells that could be suitable; a single cell (even low probability) moves a species from "gap" to non-gap status.  IUCN more clustered by polygon boundaries; so no scattered cells to accidentally fall into an MPA. 
    - introducing errors: AM = likely commission (overestimates species with tiny bits of range in MPAs) vs IUCN = possible omission (misses possible real habitat that falls within MPAs)
- hypothesis to explain more 10%+ species - ten random samples most have IUCN polygons just east of Australia where a large MPA is indicated - Great Barrier Reef I assume - a single large poly falling in GBR gets a high score, while for the same species, many of the AM ranges (also in same area) scatter suitable area over larger extent, so lower percent of species range is protected
    - introducing errors: IUCN = likely omission (skipping possible real habitat outside of MPAs) and likely commission (possibly overcounting real habitat in MPAs e.g. GBR); AquaMaps = probably a little commission (possibly underestimating, if it overpredicts lots of range outside MPAs, thus reducing the proportion of protected range)

@@

AquaMaps and IUCN are two of the most important databases providing species distribution information across a wide range of marine species, taxa, and regions.  The different methodologies behind the creation of each database is likely to introduce different rates of commission and omission errors, which can have a profound impact on our understanding of species ranges.  By comparing overlapping species maps from these two databases, we have identified likely sources of error, allowing data users to address or account for them.  For users of AquaMaps, IUCN, and other datasets of species distribution, reducing the likelihood of these errors is critical to effective biodiversity research and conservation action.
